A thermodynamic view of dusty protoplanetary disks I: axisymmetric stability

### 2016-11-18 model.tex
home/work sync
edits on the plane

For the disk problem the external potential $\Phi$ is  
that for a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant, and $(r,\phi, z)$ are
cylindrical co-ordinates centered on the star. 
The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain analytic expressions for the disk equilibria.

@@

We can compare these sources by 
evaluating them using the equilibrium
solutions in \S\ref{steady_state}. We assume the disk is radially
smooth so that $\p_r\sim 1/r$, and the dust-to-gas ratio
$\epsilon\ll1$. This gives

\begin{align}\label{vshear_split}
  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
 % \epsilon \frac{\mathrm{max}\left(\delta^2,
  %  1\right)}{\left|q\right|\left(\epsilon + \delta^2\right)},
 \epsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(1+\epsilon\right)\delta^2},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio dominates 
over that due to the radial temperature gradient for thin dust layers
such that $\delta^2\ll \epsilon$. Otherwise, vertical shear is
associated with $\p_rc_s^2$.

@@

The physical origin for this vertical buoyancy, which exists even in vertically isothermal
disk, lies in the fact that coupling gas to dust particles increases
the gas' inertia, but pressure (i.e. restoring) forces are
unchanged. The increased weight of the fluid resists vertical
oscillations and hence there is an associated buoyancy force. However,
if $\tstop\neq0$ so the gas-dust coupling is imperfect, then gas is no
longer `weighed down' by the dust, and can slip past it. This
diminishes the dust-induced buoyancy.

@@

In this section we discuss the basic properties of the 
dusty-gas mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $f$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta f(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Thus perturbations have time dependence $e^{st +
  \ii\omega t}$.

@@

Positive work is done by a fluid parcel if $-\omega \imag\left(\Delta  
P\Delta\rho^*\right)>0$. Without loss of generality, take $\omega>0$
and consider a mass element with $\Delta\rho = 1$. Then positive work 
requires $\imag\left(\Delta
P\right)<0$.  This corresponds to pressure perturbations \emph{lagging}
behind that in density (or displacements), since the phase of
the assumed perturbations increase in time.

%$\imag\left(\Delta\rho\right)=0$.

@@

For dusty gas, the work done during oscillations 
 is attributed to finite dust-gas drag. The relative
drift between gas and dust causes a phase difference between the two
components, and hence between pressure and total density.

>>>>

For the disk problem the external potential $\Phi$ is  
that for a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right), 
\end{align}
where $G$ is the gravitational constant, and $(r,\phi, z)$ are
cylindrical co-ordinates centered on the star. 
The second equality is the 
thin-disk approximation for the disk potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria.

@@

We can compare these sources by 
evaluating them using the equilibrium
solutions in \S\ref{steady_state}. We assume the disk is radially
smooth so that $\p_r\sim 1/r$. % and the dust-to-gas ratio
% $\epsilon\ll1$. 
This gives

\begin{align}\label{vshear_split}
  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
 % \epsilon \frac{\mathrm{max}\left(\delta^2,
  %  1\right)}{\left|q\right|\left(\epsilon + \delta^2\right)},
 \epsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(1+\epsilon\right)\delta^2},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio dominates 
over that due to the radial temperature gradient for thin dust layers
such that $\delta^2\ll \epsilon$. Otherwise, vertical shear is
associated with the radial temperature gradient. %However, axisymmtric
%\emph{instability} always caused by $\p_rc_s^2$.

@@

The physical origin for this vertical buoyancy, which exists even in vertically isothermal
disks, is that coupling gas to dust particles increases
the fluid's inertia, but pressure (i.e. restoring) forces are
unchanged. The increased weight of the fluid resists vertical
oscillations and hence there is an associated buoyancy force. However,
if $\tstop\neq0$ so the gas-dust coupling is imperfect, then gas is no
longer `weighed down' by the dust, and can slip past it. Thus finite
drag diminishes the dust-induced buoyancy.

@@

In this section we discuss the basic properties of the 
dusty-gas mixture using Eqs.\ref{masseq}--\ref{momeq} with 
Eq. \ref{eff_energy} in place of the dust continuity equation. Our
discussion is based on the corresponding axisymmetric linearized
equations. A variable $X$ is
subject to Eulerian perturbations of the form 
\begin{align}
 \real\left[ \delta X(r,z)\exp{\left(-\ii\sigma t\right)}\right], 
\end{align}
where $\sigma$ is the complex mode frequency. For later convenience we
write 
\begin{align}
  \sigma = \ii s - \omega,
\end{align}
where $s$ and $\omega$ are the growth rate and real frequencies,
respectively. Thus perturbations have time dependence $e^{st +
  \ii\omega t}$.

@@

Positive work is done by a fluid parcel if $-\omega \imag\left(\Delta  
P\Delta\rho^*\right)>0$. Without loss of generality, take $\omega>0$
and consider a mass element with $\Delta\rho = 1$. Then positive work 
requires $\imag\left(\Delta
P\right)<0$.  This corresponds to pressure perturbations \emph{lagging}
behind that in density (or displacements).%, since the phase of
%the assumed perturbations increase in time.  
%$\imag\left(\Delta\rho\right)=0$.

@@

The discussion above applies to any single fluid with a pressure and
density. In the case of interest --- dusty gas --- the work done 
is attributed to finite dust-gas drag. 
%For dusty gas, the work done during oscillations 
 %is attributed to finite dust-gas drag. 
The relative 
drift between gas and dust causes a phase difference between the two
components, and hence between pressure and total density.

### 2016-11-18 paper.tex
edits on the plane

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks I:
axisymmetric stability}  
\author{Min-Kai Lin\altaffilmark{1,2,3}}
\author{Andrew N. Youdin\altaffilmark{1}}
%\affil{Department of Astronomy and Steward Observatory, 
%University of
 % Arizona, Tucson, AZ 85721, USA\\
%Institute of Astronomy and Astrophysics, Academia Sinica, 
%Taipei 10617, Taiwan}
\altaffiltext{1}{Department of Astronomy and Steward Observatory, 
University of
Arizona, Tucson, AZ 85721, USA}
\altaffiltext{2}{Steward Theory Fellow}
\altaffiltext{3}{Institute of Astronomy and Astrophysics, Academia Sinica, 
Taipei 10617, Taiwan
}
\email{mklin@asiaa.sinica.edu.tw}

@@

Under these approximations, we show that the equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas, which is is analogous to the advection
of entropy following an adiabatic fluid. We will find that the   
effective entropy of isothermal dusty gas scales with its gas
fraction. Finite dust-gas drag act as effective heating/cooling: the relative
drift between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

@@

By re-casting the dusty problem into identical form as standard  
hydrodynamics, we can apply well-known methods to generalize previous stability 
analyses for pure gas disks to the dusty case. %We explicitly demonstrate this for
%perfectly coupled dust in strictly and locally isothermal disks. 
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally
stable to axisymmetric perturbations, regardless of how  thin the dust
layer is. We put forward a thermodynamical thermodynamical interpretation of
dust-gas drag instabilities, and apply it to the
streaming instability \citep[SI, ][]{youdin05a,youdin07b,jacquet11}. 
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}.

@@

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature is time-independent but varies with radius
\citep{chiang97}. Then the pressure of a fluid parcel varies with 
position as well. There is no reason to expect $P$ and $\rho$ to
be in phase. In fact, we will show that this is responsible for 
the vertical shear instability \citep{lin15}.

>>>>

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks I:
axisymmetric stability}
 % \title{Dust-free modeling of dusty protoplanetary disks I:
%axisymmetric stability}
\author{Min-Kai Lin\altaffilmark{1,2,3}}
\author{Andrew N. Youdin\altaffilmark{1}}
%\affil{Department of Astronomy and Steward Observatory, 
%University of
 % Arizona, Tucson, AZ 85721, USA\\
%Institute of Astronomy and Astrophysics, Academia Sinica, 
%Taipei 10617, Taiwan}
\altaffiltext{1}{Department of Astronomy and Steward Observatory, 
University of
Arizona, Tucson, AZ 85721, USA}
\altaffiltext{2}{Steward Theory Fellow}
\altaffiltext{3}{Institute of Astronomy and Astrophysics, Academia Sinica, 
Taipei 10617, Taiwan
}
\email{mklin@asiaa.sinica.edu.tw}

@@

Under these approximations, we show that the equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas, which is analogous to the advection
of entropy following an adiabatic fluid. We will find that the   
effective entropy of isothermal dusty gas scales with its gas
fraction. Finite dust-gas drag act as effective heating/cooling: the relative
drift between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.

@@

By re-casting the dusty problem into identical form as standard  
hydrodynamics, we can apply well-known methods to generalize previous stability 
analyses for pure gas disks to the dusty case. %We explicitly demonstrate this for
%perfectly coupled dust in strictly and locally isothermal disks. 
We show that strictly isothermal protoplanetary disks with
perfectly-coupled dust are generally
stable to axisymmetric perturbations, regardless of how  thin the dust
layer is. We put forward a thermodynamical interpretation of
dust-gas drag instabilities, and apply it to the
streaming instability \citep[SI, ][]{youdin05a,youdin07b,jacquet11}. 
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}.

@@

There are several situations where the pressure and density of a fluid
are not in phase. The obvious case is if the fluid is subject to
external heating/cooling. For example, in strongly irradiated
protoplanetary disks the
disk temperature  $T(r)$ is time-independent but varies with radius
\citep{chiang97}. Since $P\propto \rho T$ for an ideal gas, the pressure of a fluid parcel varies with 
position as well. There is no reason to expect $P$ and $\rho$ to
be in phase. In fact, we will show that this is responsible for 
the vertical shear instability \citep{lin15}.

### 2016-11-18 summary.tex
tiny edits
edits on the plane

\subsection{Generalization to locally polytropic disks}

%\subsubsection{Locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
adiabatic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  \seff = \ln{\left[K^{1/\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

@@

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas with minimal modification. In fact, no
modification is needed  
for strictly isothermal or polytropic gas perfectly-coupled to dust:  
the standard adiabatic energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15}, which is also
featured in many hydrodynamics codes.

>>>>

\subsection{Generalization to locally polytropic disks}\label{gen_poly}

%\subsubsection{Locally polytropic disks}
We can extend the dusty/adiabatic gas correspondence to 
other fixed equations of state. As an example, consider the locally
polytropic disk 
\begin{align}
  P = K(r,z)\rhog^{\Gamma}, 
\end{align}
where $K$ is a prescribed function and $\Gamma$ is the constant
polytropic index. Then eliminating $\tepsilon$ from the dust equation
\ref{dusteq} gives 
\begin{align}\label{poly_energy}
  \frac{D P}{D t} &= - \Gamma P\nabla\cdot\bm{v}  + P \bm{v}\cdot\nabla
  \ln{K} + \mathcal{C},\\
  \mathcal{C}& = \frac{\Gamma P}{\rhog}\nabla\cdot\left(\tepsilon\tstop\nabla
  P\right).
\end{align}
Thus the dusty gas behaves like a pure gas with adiabatic index
$\Gamma$. The entropy is given by 
\begin{align}
  \seff = \ln{\left[K^{1/\Gamma}(r,z)\left(1 - \tepsilon\right)\right]}.  
\end{align}
The results of \S\ref{iso_perfect}---\ref{dust_work} remain valid for the strictly 
polytropic disk with $K=$constant, while one sets $c_s^2\to K$ in
\S\ref{dusty_vsi_int}. 
%When $K$ is constant, the corresponding Solberg-Hoiland criteria for
%axisymmetric stability may be obtained from that in standard adiabatic 
%hydrodynamics \citep[e.g.][]{tassoul78} by
%using this definition of entropy in
%Eqs. \ref{dusty_solberg1}---\ref{dusty_solberg2}.

@@

\subsection{Implications for numerical simulations}
The dusty/adiabatic gas equivalence means that pure
gas dynamics codes can be applied to simulate locally
isothermal/polytropic dusty gas. One need not add any new 
equations to evolve the dust component: 
%In fact, no
%modification is needed  
%for strictly isothermal or polytropic gas perfectly-coupled to dust:  
 the standard energy equation substitutes for the dust continuity
equation. When $c_s$ or $K$ is not constant and/or $\tstop\neq0$ one
should add corresponding source terms in the energy equation
(e.g. Eq. \ref{poly_energy}). 
The source term associated with dust-gas drag, $\mathcal{C}$, is 
analogous to radiative diffusion \citep{price15}, which is also
featured in many hydrodynamics codes.
